---
title: "Comparative Evaluation of Microbiome Simulation Methods Using Real Dietswap Data"
subtitle: "Course: Microbiome Data Analysis — Assignment 1"
author: "Maryam Shojaei Shahrokhabadi"
output:
  pdf_document:
    latex_engine: xelatex
    toc: false
  word_document:
    toc: false
date: "`r Sys.Date()`"
header-includes:
- \usepackage{ragged2e}
- \raggedright
- \usepackage{booktabs}
- \usepackage{longtable}
---

```{r libraries, include=FALSE}
library(dplyr)
library(patchwork)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(microbiome)
library(ggridges)
library(stringr)
library(viridis)
```

## 1. Background

Recent advances in sequencing technologies have greatly improved our understanding of the human microbiome and its links to a wide range of diseases and biological traits. In particular, modern methods such as 16S rRNA sequencing enable rapid, parallel assessment of the relative abundances of all taxa across entire microbial communities. Although microbiome data are increasingly available, statistical analysis remains challenging. Microbiome datasets have unique characteristics that are difficult to model analytically, including sparsity (most taxa are absent in a given sample), overdispersion (the variance of read counts exceeds that assumed by standard parametric models), compositionality (the read counts within a sample sum to a constant), and strong correlations between taxa. As microbiome data have grown, a wide range of statistical methods has emerged to study these relationships. This rapid development has created a strong demand for straightforward, efficient, and reliable tools that can generate realistic microbiome simulations for rigorous methodological evaluation.

Traditional simulation approaches for microbiome data often rely on simplified assumptions that fail to reflect the complex multivariate structure observed in real datasets. Some methods capture marginal distributions but not inter-taxa correlations, while others preserve correlations but struggle with zero inflation, overdispersion, or compositionality.

MIDASim (Microbiome Data Simulator) is designed to overcome these limitations, by simultaneously modelling both the presence–absence patterns and abundance structures of a template microbiome dataset. However, studies evaluating the performance of MIDASim remain limited. The objective of this project is to systematically assess the performance of MIDASim in generating realistic microbiome data and to evaluate how closely the simulated data resemble the original template dataset. The evaluation will proceed in three distinct phases. Assessment of performace, assessment of consistency and robustness, and a comparison with other methods. 

## 2. Methods

## 2.1 Template Data and Parametric Simulation (MIDASim)
We used MIDASim to generate realistic microbiome datasets by adapting an existing template dataset. The template was derived from the Dietswap study by O’Keefe et al (2015), in which 22 African American (AAM) participants were placed on a high-fibre, low-fat African-style diet and 16 rural South African (AFR) participants were placed on a high-fat, low-fibre Western-style diet under controlled conditions. For this analysis, we focused on samples collected at day 2, resulting in a dataset containing 130 taxa across 37 samples.

To generate synthetic datasets, we employed the parametric mode of MIDASim, which implements a two-step procedure: first modeling the presence-absence of taxa, and then modeling the non-zero relative abundances. In this framework, MIDASim first extracts taxon-specific characteristics from the template dataset—including mean relative abundances, marginal probabilities of zeros, dispersion estimates, and inter-taxa correlation structures—and fits a generalized gamma model to the non-zero values. The model parameters are used to preserve the structure of the original microbiome community, including the relationships between taxa and the distribution of abundances, in the simulated datasets. To control the magnitude and structure of group-level differences, we set the effect size to 1.2 and designated 10% of taxa as causal. Additionally, 70% of the taxa were randomly assigned to one subset and the remaining 30% to another, with their relative abundances adjusted in opposite directions to introduce bidirectional, group-specific shifts.

## 2.2 Comparative Simulation Methods
To assess the performance of the MIDASim parametric approach, we generated comparable synthetic datasets using three alternative methods: MIDASim's nonparametric mode, SparseDOSSA2, and SPsimSeq.

### A. MIDASim Nonparametric Mode 
The nonparametric mode generates data directly by drawing from the empirical distribution of the template dataset. Similar to the parametric mode, the first step models the presence–absence of taxa. For taxa determined to be non-zero in this step, a rank-based approach is used to assign their non-zero relative abundances.

### B. SparseDOSSA2
SparseDOSSA2 is a simulation method for generating realistic synthetic microbiome datasets by learning key characteristics from a template dataset. The method models microbial abundances using zero-inflated log-normal marginal distributions and links features via a Gaussian copula to preserve inter-feature correlation structures. Model parameters are estimated using a penalized expectation-maximization (EM) algorithm, including per-feature fraction of non-zero abundances, the mean and standard deviation of non-zero abundances, and feature-feature correlations.

### C. SPsimSeq
SPsimSeq is a semi-parametric simulator originally developed for RNA-seq data but widely applied to microbiome datasets. It models count data using a negative binomial distribution to capture the high mean–variance relationship typical of sequencing data. The method separately models zero inflation to account for high sparsity and uses a gaussian copula to preserve correlations among taxa.

## 2.3 Assessment and Comparison
To assess the performance of MIDASim, a single simulated dataset was used. For this dataset, key taxon-level metrics—including mean abundance, variance, coefficient of variation (CV), and fraction of zeros—were calculated. These metrics capture essential microbiome characteristics: central tendency (mean), dispersion (variance and CV), and sparsity (presence–absence patterns). Relationships between these properties in the simulated and real datasets were visualized separately for the AAM and AFR groups, including:

- Mean Abundance vs. Variance
- Mean Abundance vs. CV
- Mean Abundance vs. Fraction of Zeros


To assess the consistency of MIDASim, we generated 100 independent simulated datasets. For each simulated dataset, key statistical metrics, including the mean, variance, CV, and fraction of zeros, will be calculated for each taxon. To quantify deviations, mean abundances were divided into bins, and within each bin, the difference between simulated and real datasets was computed for each metric. This process yielded 100 difference values per bin for each metric. The distributions of these differences will be visualized using boxplots or ridge plots to display the full distribution within each interval. The spread of the boxplots reflects the consistency of MIDASim simulations: smaller variance across boxplots indicates higher consistency and better reproducibility of real data characteristics across multiple simulated datasets.

Finally, the parametric version of MIDASim was compared with (a) MIDASim’s nonparametric mode, (b) SparseDOSSA2, and (c) SPsimSeq to evaluate which method best captures the key statistical properties of microbiome data, including mean, variance, CV, and fraction of zeros.

```{r setup_one_dataset, include=FALSE}
## loading all data
load("C:/Users/Administrator/Desktop/Uhasselt/microbiom/Setting6_n200_b5_rnt13.RData")
phyloseq_obj <- Setting6_n200_b5_rnt13[[1]]

data("dietswap")

load("C:/Users/Administrator/Desktop/Uhasselt/microbiom/nonparametric_1.RData") 
load("C:/Users/Administrator/Desktop/Uhasselt/microbiom/SparseDOSSA2_simulated.RData")
load("C:/Users/Administrator/Desktop/Uhasselt/microbiom/Spsimseq_simulated.RData") 

## helper function
get_stats <- function(otu, samples) {
  otu_sub <- otu[samples, , drop = FALSE]
  lib_sizes <- rowSums(otu_sub)
  otu_cpm <- sweep(otu_sub, 1, lib_sizes / 1e6, FUN = "/")
  
  mean_cpm <- colMeans(otu_cpm)
  var_cpm  <- apply(otu_cpm, 2, var)
  cv_cpm   <- sqrt(var_cpm) / (mean_cpm + 1e-8)  
  frac_zero <- colMeans(otu_sub == 0)
  
  list(
    log_mean  = log2(mean_cpm + 1),
    log_var   = log2(var_cpm + 1),
    log_cv    = log2(cv_cpm + 1),
    frac_zero = frac_zero
  )
}

## Real data
otu_real <- as(otu_table(dietswap), "matrix")
if (taxa_are_rows(dietswap)) {
  otu_real <- t(otu_real)
}
meta_real <- sample_data(dietswap)
groups_real <- unique(meta_real$nationality)

## for parametric MIDASIM
otu1 <- as(otu_table(phyloseq_obj), "matrix")
if (taxa_are_rows(phyloseq_obj)) {
  otu1 <- t(otu1)
}
meta1 <- sample_data(phyloseq_obj)
groups1 <- unique(meta1$group)

## non-parametric Midasim
otu_sim <- as(otu_table(ps_sim), "matrix")
if (taxa_are_rows(ps_sim)) {
  otu_sim <- t(otu_sim)
}
meta_mid_nonpar <- sample_data(ps_sim)
groups_mid_nonpar <- unique(meta_mid_nonpar$Group)

## --- FILTER REAL DATA BASED ON SIMULATED TAXA ---
sim_taxa <- colnames(otu_sim)   
otu_real_non <- otu_real[, sim_taxa, drop = FALSE]

## simulated data (SPsimSeq)
otu_spsim <- t(sim_micro[[1]]$counts)   # samples × taxa
meta_spsim <- sim_micro[[1]]$colData
groups_spsim <- unique(meta_spsim$Group)

## simulated data (SparseDOSSA2)
otu_sim2 <- as(otu_table(ps_sim2), "matrix")
if (taxa_are_rows(ps_sim2)) {
  otu_sim2 <- t(otu_sim2)
}
meta_Spars<- sample_data(ps_sim2)
groups_Spars <- unique(meta_Spars$Group)
sim_taxa <- taxa_names(ps_sim2)  
length(sim_taxa)
# Real data
otu_real <- as(otu_table(dietswap), "matrix")
if (taxa_are_rows(dietswap)) {
  otu_real <- t(otu_real)
}
otu_real_filtered <- otu_real[, sim_taxa, drop = FALSE]  # keep only SparseDOSSA2 taxa

```

## 3. Results

The Dietswap dataset comprised 130 taxa (OTUs) across 37 samples, with approximately 21% zeros, indicating moderate overall sparsity. Library sizes were highly skewed across samples and the CV of the data was 5.13, suggesting substantial overdispersion. At the taxon level, CV values ranged from 0.53 to 14.90, with a median of 1.36 and a mean of 2.55.

## 3.1.	Assessment of MIDASim Performance

Visual inspection of the relationships between mean abundance and variance and mean abundance and CV showed that the simulated data closely resembled the real data across both AAM and AFR groups (Figures 1a–1b). For the relationship between mean abundance and the fraction of zeros (sparsity), the simulation correctly reproduced the trend that low-abundance taxa were more likely to be absent. However, it deviated from the real data for taxa with intermediate abundances: in the simulations, the fraction of zeros exhibited greater variability and did not follow the distribution observed in the real dataset (Figure 1c).

To formally evaluate whether the simulated data reproduced the mean–variance relationship observed in the real dataset, we modeled the association between log-mean abundance and log-variance using a linear regression with an interaction term between mean abundance and data source (real vs. simulated). Because the mean–variance trend appeared approximately linear on the log–log scale, a linear model was appropriate:
\[
\log\,(\text{var})
= \beta_{0}
\;+\; \beta_{1} \times \log\,(\text{mean})
\;+\; \beta_{2} \times \text{source}
\;+\; \beta_{3} \times \bigl(\log\,(\text{mean}) \times \text{source}\bigr)
\]

Where `source` indicates whether the observation comes from **Real Data** or **Simulated Data**, and \(\beta_{3}\) represents the **difference in slopes** between the simulated and real datasets. When \(\beta_{3} = 0\): the simulator **perfectly matches** the slope observed in the real data, whereas values of \(\beta_{3} \neq 0\): the simulator **deviates** from the real-data slope.

The estimated slope difference was not significant in the AAM group, indicating that the simulated data closely reproduced the mean–variance structure of the real samples. In the AFR group, the slope difference was statistically significant but quantitatively small relative to the overall slope of the mean–variance relationship ($\beta_3$ = 0.09, SE = 0.04, p = 0.03), suggesting that although a detectable difference existed, the deviation between the real and simulated data was modest.

```{r setting ploting one sim, echo=FALSE, warning = FALSE, message = FALSE}

## plot MIDASIM parametric
plots_var <- list()
plots_zero <- list()
plots_cv <- list()
interaction_tables <- list()

for (i in 1:2) {
  samples1 <- rownames(meta1)[meta1$group == groups1[i]]
  st1 <- as.data.frame(get_stats(otu1, samples1)) %>%
    mutate(source = paste0("Simulated Data"))
  
  samples2 <- rownames(meta_real)[meta_real$nationality == groups_real[i]]
  st2 <- as.data.frame(get_stats(otu_real, samples2)) %>%
    mutate(source = paste0("Real Data"))
  
  df <- bind_rows(st1, st2)
  df1 <- df %>% filter(source == "Real Data")
  df2 <- df %>% filter(source == "Simulated Data")
  lm_real  <- lm(log_var ~ log_mean, data = df1)
  lm_sim   <- lm(log_var ~ log_mean, data = df2)
  df$source <- factor(df$source, levels = c("Real Data", "Simulated Data"))
  lm_interaction <- lm(log_var ~ log_mean * source, data = df)

  group_title <- paste("Comparision within nationality", groups_real[i])
  
  # --- Plot 1: variance vs mean ---
  p <- ggplot(df, aes(x = log_mean, y = log_var, color = source)) +
    geom_point(alpha = 1, size = 1) +
    geom_smooth(aes(group = source), method = "lm", se = FALSE, linetype = "dashed") +
    labs(
      x = "mean log2-CPM",
      y = "var log2-CPM",
      title = group_title
    ) +
    theme_minimal() +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0, size = 8),
      axis.title = element_text(size = 8) 
    )
  
  # --- Plot 2: zeros vs mean ---
  p_zero <- ggplot(df, aes(x = log_mean, y = frac_zero, color = source)) +
    geom_point(alpha = 0.7, size = 1) +
    labs(
      x = "mean log2-CPM",
      y = "fraction of zeros",
      title = group_title
    ) +
    theme_minimal() +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0, size = 8),
      axis.title = element_text(size = 8)
    )
  
  # --- Plot 3: cv vs mean ---
  p_cv <- ggplot(df, aes(x = log_mean, y = log_cv , color = source)) +
    geom_point(alpha = 0.7, size = 1) +
    labs(
      x = "mean log2-CPM",
      y = "log2-CV",
      title = group_title
    ) +
    theme_minimal() +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0, size = 8),
      axis.title = element_text(size = 8)
    )
  
  plots_var[[i]] <- p
  plots_zero[[i]] <- p_zero
  plots_cv[[i]] <- p_cv
  
# --- Extract coefficients ---
coef_tab <- coef(summary(lm_interaction))

# Specify which terms you want in the table
terms_to_include <- c( "log_mean", "log_mean:sourceSimulated Data")

# Subset the coefficient table
coef_subset <- coef_tab[terms_to_include, , drop = FALSE]

# Create clean table with readable names
interaction_table <- data.frame(
  Nationality = groups_real[i],
  Term = c(
    "\u03B2\u2081 Main effects",
    "\u03B2\u2083   Slope difference (Simulated vs Real)"
  ),
  Estimate = round(coef_subset[, "Estimate"], 2),
  `Std. Error` = round(coef_subset[, "Std. Error"], 2),
  `t value` = round(coef_subset[, "t value"], 2),
  `p-value` = round(coef_subset[, "Pr(>|t|)"], 3),
  check.names = FALSE
)

# Store table in the list
interaction_tables[[i]] <- interaction_table
}
```

```{r table, echo=FALSE, warning = FALSE, message = FALSE}

# Combine all interaction tables
interaction_tables_combined <- bind_rows(interaction_tables)

# Clean up and add separate Beta column
interaction_tables_clean <- interaction_tables_combined %>%
  mutate(
    Beta = gsub(" .*", "", Term),           # Extract β₀, β₁, β₃
    Term = gsub("^β[0-9]+ ", "", Term),    # Remove beta from Term
    Nationality = as.character(Nationality) # Ensure clean display
  ) %>%
  select(Beta, Nationality, Term, Estimate, `Std. Error`, `t value`, `p-value`)

rownames(interaction_tables_combined) <- NULL

# Revised cleaning and selection
interaction_tables_combined_cleaned <- interaction_tables_combined %>%
mutate(
  # 1. Clean up the 'Term' column to remove the descriptive text
  Clean_Term = case_when(
    grepl("log_mean", Term) ~ "Main effects",
    grepl("Slope difference", Term) ~ "  Slope difference (Simulated vs Real)",
    # Handle other terms if they exist
    TRUE ~ Term
  )
) %>%
# Select and reorder the final columns
select( Nationality, Term = Clean_Term, Estimate, `Std. Error`, `t value`, `p-value`)


# Print table
kable(interaction_tables_combined_cleaned,
caption = "Slope differences and main effects for each nationality",
booktabs = TRUE) %>%
kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) %>%
row_spec(0, bold = TRUE)

```

```{r ploting one simulated data, echo=FALSE, warning = FALSE, message = FALSE}
plot1_midasim <- plots_var[[1]]
plot2_midasim <- plots_var[[2]]
plot3_midasim <- plots_zero[[1]]
plot4_midasim <- plots_zero[[2]]
plot5_midasim <- plots_cv[[1]]
plot6_midasim <- plots_cv[[2]]

#  Variance plots (AAM + AFR side by side)
  plot_var_combined <- (plot1_midasim / plot2_midasim) +
  plot_layout(ncol = 1) +
  plot_annotation(
    caption = paste0(
      "Figure 1a: Comparison of the variance versus mean relationship for taxon abundances \n",
      "between the real Dietswap dataset and a single simulated dataset generated using MIDASim (parametric mode).\n",
      "Results are shown separately for the African American (AAM) and Rural South African (AFR) groups."
    ),
    theme = theme(
      plot.caption = element_text(size = 9, hjust = 0, lineheight = 1.2)
    )
  )

#  Zero-fraction plots
plot_zero_combined <- plot3_midasim / plot4_midasim +
  plot_layout(ncol = 1)+
  plot_annotation(
    caption = paste0(
      "Figure 1b: Comparison of the fraction of zeros versus mean relationship for taxon abundances\n",
      "between the real Dietswap dataset and a single simulated dataset generated using MIDASim (parametric mode).\n",
      "Results are shown separately for the African American (AAM) and Rural South African (AFR) groups."
    ),
    theme = theme(
      plot.caption = element_text(size = 9, hjust = 0, lineheight = 1.2)
    )
  )


#  CV plots
plot_cv_combined <- plot5_midasim / plot6_midasim +
  plot_layout(ncol = 1)+
  plot_annotation(
    caption = paste0(
      "Figure 1c: Comparison of the coefficient of variation (CV) versus mean relationship for taxon abundances\n",
      "between the real Dietswap dataset and a single simulated dataset generated using MIDASim (parametric mode).\n",
      "Results are shown separately for the African American (AAM) and Rural South African (AFR) groups."
    ),
    theme = theme(
      plot.caption = element_text(size = 9, hjust = 0, lineheight = 1.2)
    )
  )

# Display them
plot_var_combined
```
```{r, fig.height=4, fig.width=6, out.height="45%", out.width="90%", echo=FALSE, warning = FALSE, message = FALSE}
plot_zero_combined
plot_cv_combined
```

```{r setup_100_simulation, include=FALSE}
#--------real data ------
otu_real <- as(otu_table(dietswap), "matrix")
if (taxa_are_rows(dietswap)) {
  otu_real <- t(otu_real)
}
meta_real <- sample_data(dietswap)
groups_real <- unique(meta_real$nationality)

#--------loading 100 simulated data ------
setwd("C:/Users/Administrator/Desktop/Uhasselt/microbiom/Setting6(n_200_b_5_rnt_13)")

files <- list.files(pattern = "^sim[0-9]+\\.Rdata$")

sim_list <- lapply(files, function(f) {
  e <- new.env()
  load(f, envir = e)
  as.list(e)
})

#--------helper function to calculate mean, Var, CV ------
get_stats <- function(otu, samples) {
  otu_sub <- otu[samples, , drop = FALSE]
  lib_sizes <- rowSums(otu_sub)
  otu_cpm <- sweep(otu_sub, 1, lib_sizes / 1e6, FUN = "/")
  
  mean_cpm <- colMeans(otu_cpm)
  var_cpm  <- apply(otu_cpm, 2, var)
  cv_cpm   <- sqrt(var_cpm) / (mean_cpm + 1e-8)  
  frac_zero <- colMeans(otu_sub == 0)
  
  list(
    log_mean  = log2(mean_cpm + 1),
    log_var   = log2(var_cpm + 1),
    log_cv    = log2(cv_cpm + 1),
    frac_zero = frac_zero
  )
}

#--------Creating phyloseq objects from the simulated counts------
create_phyloseq <- function(midas_data) {
  # otu table
  midas_otutable <- midas_data[[2]]
  rownames(midas_otutable) <- paste0("sample", 1:nrow(midas_otutable))  # Set row names for samples
  
  # taxa table
  n.taxa <- ncol(midas_otutable)
  midas_taxtable <- matrix(nrow = n.taxa, ncol = 2)
  midas_taxtable[, 1:2] <- c(midas_data[[1]], midas_data[[3]])
  colnames(midas_taxtable) <- c("DA_ind", "cens.prob")
  midas_taxtable <- as.data.frame(midas_taxtable) 
  midas_taxtable$isDA <- ifelse(midas_taxtable$DA_ind == 1, "TRUE", "FALSE")
  rownames(midas_taxtable) <- colnames(midas_otutable)
  midas_taxtable <- as.matrix(midas_taxtable)
  
  # sample data
  midas_samdata <- as.data.frame(midas_data[[4]])
  rownames(midas_samdata) <- rownames(midas_otutable)
  colnames(midas_samdata) <- "x.index"
  midas_samdata$group <- ifelse(midas_samdata$x.index == 0.5, 1, 0)  # Create a 'group' column for the groupings
  
  # Create phyloseq object
  midas_phyloseq <- phyloseq(otu_table(midas_otutable, taxa_are_rows = FALSE),
                             sample_data(midas_samdata),
                             tax_table(midas_taxtable))
  
  return(midas_phyloseq)
}


physeq_list <- vector("list", length(sim_list))

# Loop over all simulations
for (i in seq_along(sim_list)) {
  physeq_list[[i]] <- create_phyloseq(sim_list[[i]][["midas.data"]])
}

#--------calculate statistics for 100 simulated------
all_stats <- lapply(physeq_list, function(physeq_obj) {
  otu_mat <- as(otu_table(physeq_obj), "matrix")
  if (taxa_are_rows(physeq_obj)) otu_mat <- t(otu_mat)
  
  meta1 <- sample_data(physeq_obj)
  groups1 <- unique(meta1$group)
  
  # For each group(AAF,AFM), calculate stats
  group_stats <- lapply(groups1, function(g) {
    samples_g <- rownames(meta1)[meta1$group == g]
    get_stats(otu_mat, samples_g)
  })
  
  group_stats
})

#--------calculate statistics for Real data------
all_stats_real <- lapply(groups_real, function(g) {
  
  samples_g <- rownames(meta_real)[meta_real$nationality == g]
  
  
  get_stats(otu_real, samples_g)
})


#--------fixing intervals(using real data)------
log_mean_real <- all_stats_real[[1]]$log_mean
log_var_real  <- all_stats_real[[1]]$log_var

############## function for intervals ###########
compute_diff_per_interval <- function(
    all_stats_real,
    all_stats,
    metric = c("log_var", "log_cv", "frac_zero"),
    num_intervals = 10,
    group_index = 1,
    breaks = NULL      # << NEW
) {
  metric <- match.arg(metric)
  
  # Extract real values
  log_mean_real <- all_stats_real[[group_index]]$log_mean
  metric_real   <- all_stats_real[[group_index]][[metric]]
  
  # 1. Quantile breaks (use shared breaks if provided)
  if (is.null(breaks)) {
    percentiles <- seq(0, 1, length.out = num_intervals + 1)
    breaks <- unique(quantile(log_mean_real, probs = percentiles, type = 7))
  }
  
  # 2. Real intervals
  intervals_real <- cut(log_mean_real, breaks = breaks,
                        include.lowest = TRUE, dig.lab = 4)
  
  # 3. Mean of chosen metric for real data
  real_mean_by_interval <- tapply(metric_real, intervals_real, mean)
  real_mean_by_interval[is.na(real_mean_by_interval)] <- 0
  
  real_df <- data.frame(
    interval = names(real_mean_by_interval),
    mean_metric = as.numeric(real_mean_by_interval)
  )
  
  interval_levels <- levels(intervals_real)
  
  # 4. Simulated data
  sim_means <- lapply(all_stats, function(sim) {
    log_mean_i <- sim[[group_index]]$log_mean
    metric_i   <- sim[[group_index]][[metric]]
    
    intervals_i <- cut(log_mean_i, breaks = breaks,
                       include.lowest = TRUE, labels = interval_levels)
    
    mean_metric_i <- tapply(metric_i, intervals_i, mean)
    mean_metric_i[is.na(mean_metric_i)] <- 0
    
    data.frame(interval = interval_levels,
               mean_metric = as.numeric(mean_metric_i))
  })
  
  # 5. Build matrix and compute diff
  sim_matrix <- sapply(sim_means, `[[`, "mean_metric")
  diff_matrix <- sim_matrix - real_df$mean_metric
  
  K <- length(interval_levels)
  
  df_long <- as.data.frame(diff_matrix) %>%
    mutate(interval = factor(1:K)) %>%
    tidyr::pivot_longer(cols = -interval,
                        names_to = "simulation",
                        values_to = "diff") %>%
    mutate(interval_label = factor(interval, labels = interval_levels),
           metric = metric)
  
  return(df_long)
}

```

## 3.2.	Consistency Assessment

To compare the real and 100 simulated datasets across the full range of taxon abundances, we initially divided the mean abundance values into 50 narrow intervals. For each interval, we calculated the mean variance and the difference between simulated and real mean log-variance. Results for the AAM group are shown in Figure 2. The differences fluctuated widely across the 100 simulations, leading to high variability and poor reproducibility. This instability was most observed in low-abundance intervals (roughly the first 17; log-mean $\approx 0–6$) and again at very high abundances (log-mean > 15.5).

Examination of the mean-abundance distributions in both the real and simulated datasets for the AAM groups (Figure 3) showed that these narrow intervals often contained very few taxa (sometimes only 1, 2, or even none). As a result, a single simulated taxon falling into such a bin could disproportionately affect the interval’s mean variance. This scarcity of data explains the inflated variability at both ends of the abundance range.

Instead of choosing an arbitrary number of intervals (e.g., 50), we used quantile-based intervals derived from the distribution of mean abundances in the real data, ensuring that each interval contained enough taxa for stable estimates. This approach forces each interval to have roughly the same number of taxa (equal-frequency binning). We tested different minimum numbers of taxa per interval: 3 taxa per bin and 5 taxa per bin. Using 3 taxa per bin still produced unstable and noisy estimates of mean log-variance differences, even in the middle range where data are usually dense.

For 5 taxa per bin, we determined the number of intervals based on the dataset size. With 130 taxa total, approximately 26 intervals would be needed ($130 / 5\approx 26$). The real data were divided into 25 quantile-based intervals, each containing roughly the same number of taxa to avoid empty or overly sparse bins. However, because many taxa had identical mean-abundance values some percentile breakpoints were identical, which “collapsed” intervals. As a result, the final number of intervals was 24. We then calculated the mean variance, CV, and fraction of zeros for the 100 simulated datasets within these 24 intervals.

Using 5 taxa per bin, variability in mean log-variance between simulated and real data decreased noticeably, resulting in more stable estimates for both nationalities (Figures 4a). However, the simulator did not fully capture the zero-inflation observed in the real data, although the fraction of zeros was reasonably well reproduced in the tails (extreme low/high abundances) (Figures 4b), consistent with patterns seen in a single simulated dataset (Figure 1b). CV was poorly reproduced but still better than the fraction of zeros (Figures 4c). In other words, the simulator reproduced variance patterns reasonably well but struggled to replicate the strong sparsity of the real data.

# 3.3. Comparative Performance Analysis

We compared how well different simulation methods reproduce key statistical properties of the real microbiome data, focusing on three relationships: mean abundance versus variance, mean abundance versus coefficient of variation (CV), and mean abundance versus fraction of zeros.

For the non-parametric MIDASim simulations, we applied a filtering step to remove very low-abundance taxa, which resulted in 107 taxa. In contrast, SparseDOSSA2 and SPsimSeq handled rare taxa better, retaining 114 taxa. To ensure a fair comparison across all methods, the corresponding taxa in the real dataset were filtered to match those present in the simulated datasets (Figure 5a-5c).

Overall, methods that explicitly account for non-parametric distributions or semi-parametric approaches more accurately reproduced the key statistical patterns observed in the real data, particularly the relationship between mean abundance and the fraction of zeros (Figure 5b). However, they still show some limitations in fully capturing the behavior of very rare taxa.

The parametric version of MIDASim, while able to model the general structure, was less effective at capturing variability in low-abundance taxa and zero-inflation patterns. The simulated points were highly scattered in the low- to mid-abundance range (Figure 5b, panel a).


```{r plot 50 intervals(setup), echo=FALSE}

## create equal-width breaks
num_intervals <- 50
breaks <- seq(min(log_mean_real), max(log_mean_real), length.out = num_intervals + 1)
## summary(breaks)
intervals <- cut(log_mean_real, breaks = breaks, include.lowest = TRUE)
## summary(intervals)
## calculate mean of variances per interval
mean_log_var_by_interval <- tapply(log_var_real, intervals, mean)
## for sume intervals without taxa this value was NA I replaced that with 0
## to see if in simulated data in this range of mean we will have any taxa or not
mean_log_var_by_interval[is.na(mean_log_var_by_interval)] <- 0

## Convert the tapply result to a data frame
mean_log_var_df <- data.frame(
  interval = names(mean_log_var_by_interval),
  mean_log_var = as.numeric(mean_log_var_by_interval)
)

#--------calculating mean of variance for all simulated data------
mean_log_var_all_sim <- vector("list", length(all_stats))

for (i in seq_along(all_stats)) {
  
  # Extract log_mean and log_var for group 1 
  log_mean_i <- all_stats[[i]][[1]]$log_mean
  log_var_i  <- all_stats[[i]][[1]]$log_var
  
  # Assign to intervals from REAL data
  intervals_i <- cut(log_mean_i, breaks = breaks, include.lowest = TRUE)
  
  # Compute means within intervals
  mean_log_var_i <- tapply(log_var_i, intervals_i, mean)
  
  # Replace NA with 0
  mean_log_var_i[is.na(mean_log_var_i)] <- 0
  
  # Save as a data frame (same structure as mean_log_var_df)
  mean_log_var_all_sim[[i]] <- data.frame(
    interval = names(mean_log_var_i),
    mean_log_var = as.numeric(mean_log_var_i)
  )
}

# create matrix: rows = intervals, cols = simulations
sim_matrix <- matrix(NA, nrow = num_intervals, ncol = length(mean_log_var_all_sim))

for (i in seq_along(mean_log_var_all_sim)) {
  sim_matrix[, i] <- mean_log_var_all_sim[[i]]$mean_log_var
}

diff_matrix <- sim_matrix - mean_log_var_df$mean_log_var


K <- nrow(diff_matrix)

df_long <- as.data.frame(diff_matrix) %>%
  mutate(interval = factor(1:K)) %>%       # add interval index
  pivot_longer(
    cols = -interval,
    names_to = "simulation",
    values_to = "diff"
  )

## Map interval index to actual interval labels
interval_labels <- levels(intervals)  # get the interval ranges as labels
df_long$interval_label <- factor(df_long$interval, labels = interval_labels)

## create custom legend labels
custom_labels <- paste0("Interval ", seq_along(interval_labels), " = ", interval_labels)
```

```{r plot 50 intervals, echo=FALSE}
plot_50 <- ggplot(df_long, aes(x = interval_label, y = diff)) +
  geom_boxplot(outlier.size = 0.7, color = "black", fill = "grey70") +  # all boxes grey
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.title = element_text(size = 8),
    plot.caption = element_text(size = 10, face = "italic", hjust = 0.5)
  ) +
  labs(
    x = "Mean Intervals",
    y = expression("Simulated – Real ("*Delta*" mean log-variance)"),
    caption = paste0("Figure 2: Distribution of the differences between simulated and real mean log-variance\n", "across 50 mean-abundance intervals, based on 100 simulations (AAM group).")
  )


```

```{r distributions, echo=FALSE}

library(patchwork)
### plot distribution of mean_real
df_real <- data.frame(log_mean_real = log_mean_real)
p_real <- ggplot(df_real, aes(x = log_mean_real)) +
  geom_histogram(bins = 30, fill = "grey70", color = "black") +
  theme_bw() +
  labs(
    title = "(a)",
    x = "log(mean CPM)",
    y = "Count"
  )

#--------calculating mean of variance for one simulated data------
log_mean_1 <- all_stats[[1]][[1]]$log_mean
log_var_1  <- all_stats[[1]][[1]]$log_var

mean_log_var_by_interval1 <- tapply(log_var_1, intervals, mean)
# for sume intervals without taxa this value was NA I replaced that with 0
# to see if in simulated data in this range of mean we will have any taxa or not
mean_log_var_by_interval1[is.na(mean_log_var_by_interval1)] <- 0

# Convert the tapply result to a data frame
mean_log_var_df1 <- data.frame(
  interval = names(mean_log_var_by_interval1),
  mean_log_var = as.numeric(mean_log_var_by_interval1)
)

### plot distribution of mean_real
df_real1 <- data.frame(log_mean_1 = log_mean_1)
p_sim <- ggplot(df_real1, aes(x = log_mean_1)) +
  geom_histogram(bins = 30, fill = "grey70", color = "black") +
  theme_bw() +
  labs(
    title = "(b)",
    x = "log(mean CPM)",
    y = "Count"
  )

```


```{r, fig.height=4, fig.width=6, out.height="50%", out.width="90%", echo=FALSE, warning = FALSE, message = FALSE}
plot_50
```
```{r, fig.height=4, fig.width=6, out.height="40%", out.width="80%", echo=FALSE, warning = FALSE, message = FALSE}
(p_real / p_sim) +
  plot_annotation(
    caption = "Figure 3: Distribution of log-mean-CPM (a: real data, b: single simulated data)."
  ) &
  theme(
    plot.caption = element_text(size = 10, hjust = 0.5)   # centered caption
  )
```

```{r plot 25 intervals_5 variance, message=FALSE, echo=FALSE}
log_mean_all_groups <- c(
  all_stats_real[[1]]$log_mean,
  all_stats_real[[2]]$log_mean
)
num_intervals <- 25
percentiles <- seq(0, 1, length.out = num_intervals + 1)
shared_breaks <- unique(quantile(log_mean_all_groups, probs = percentiles, type = 7))

df_long_25 <- compute_diff_per_interval(
  all_stats_real, all_stats,
  metric = "log_var",
  num_intervals = 25,
  group_index = 1,
  breaks = shared_breaks
)

df_long_252 <- compute_diff_per_interval(
  all_stats_real, all_stats,
  metric = "log_var",
  num_intervals = 25,
  group_index = 2,
  breaks = shared_breaks
)

df_long_25$group  <- "AAM"
df_long_252$group <- "AFR"

df_ridge <- rbind(df_long_25, df_long_252)


brig_var <- ggplot(df_ridge, aes(x = diff, y = interval_label, fill = group)) +
  geom_density_ridges(alpha = 0.7) +
  scale_fill_viridis_d(option = "D", end = 0.9) +
  theme_ridges() +
  theme(
    # Smaller y-axis text
    axis.text.y = element_text(size = 7),       
    axis.title.y = element_text(size = 8, face = "bold", hjust = 0.5),
    
    # Smaller x-axis text
    axis.text.x = element_text(size = 8),
    axis.title.x = element_text(size = 10, face = "bold", hjust = 0.5),
    
    # Smaller legend text
    legend.title = element_text(size = 7, face = "bold"),
    legend.text  = element_text(size = 6),
    
    # Adjust strip text (facet labels)
    strip.text = element_text(size = 9, face = "bold"),
    
    # Caption formatting: smaller font and centered
    plot.caption = element_text(size = 10, hjust = 0.5, lineheight = 1.2),
    
    # Increase spacing between facets if needed
    panel.spacing = unit(1, "lines")
  ) +
  labs(
    x = expression("Simulated – Real ("*Delta*" mean log-variance)"),
    y = "Intervals",
    caption = paste0(
      "Figure 4a: Distribution of differences between simulated and real mean log-variance\n",
      "across 24 mean-abundance intervals (minimum 5 taxa per interval),\n",
      "based on 100 simulations for for both AAM and AFR."
    ),
    fill = "Group"
  ) +
  facet_grid(. ~ group)

```

```{r plot 25 intervals_5 FZ, message=FALSE, echo=FALSE}

df_long_FZ <- compute_diff_per_interval(
  all_stats_real, all_stats,
  metric = "frac_zero",
  num_intervals = 25,
  group_index = 1,
  breaks = shared_breaks
)

df_long_FZ1 <- compute_diff_per_interval(
  all_stats_real, all_stats,
  metric = "frac_zero",
  num_intervals = 25,
  group_index = 2,
  breaks = shared_breaks
)

df_long_FZ$group  <- "AAM"
df_long_FZ1$group <- "AFR"

df_ridge_FZ <- rbind(df_long_FZ, df_long_FZ1)


bridge_fz <- ggplot(df_ridge_FZ, aes(x = diff, y = interval_label, fill = group)) +
  geom_density_ridges(alpha = 0.7) +
  scale_fill_viridis_d(option = "D", end = 0.9) +
  theme_ridges() +
  theme(
    # Smaller y-axis text
    axis.text.y = element_text(size = 7),       
    axis.title.y = element_text(size = 8, face = "bold", hjust = 0.5),
    
    # Smaller x-axis text
    axis.text.x = element_text(size = 8),
    axis.title.x = element_text(size = 10, face = "bold", hjust = 0.5),
    
    # Smaller legend text
    legend.title = element_text(size = 7, face = "bold"),
    legend.text  = element_text(size = 6),
    
    # Adjust strip text (facet labels)
    strip.text = element_text(size = 9, face = "bold"),
    
    # Caption formatting: smaller font and centered
    plot.caption = element_text(size = 10, hjust = 0.5, lineheight = 1.2),
    
    # Increase spacing between facets if needed
    panel.spacing = unit(1, "lines")
  ) +
  labs(
    x = expression("Simulated – Real ("*Delta*" mean log-CV)"),
    y = "Intervals",
    caption = paste0(
      "Figure 4b: Distribution of differences between simulated and real mean log-fraction of zeros\n",
      "across 24 mean-abundance intervals (minimum 5 taxa per interval),\n",
      "based on 100 simulations for for both AAM and AFR."
    ),
    fill = "Group"
  ) +
  facet_grid(. ~ group)

```

```{r, fig.height=4, fig.width=6, out.height="45%", out.width="90%", echo=FALSE, warning = FALSE, message = FALSE}
brig_var
bridge_fz
```

```{r plot 25 intervals_5 CV, message=FALSE, echo=FALSE}

df_long_cv <- compute_diff_per_interval(
  all_stats_real, all_stats,
  metric = "log_cv",
  num_intervals = 25,
  group_index = 1,
  breaks = shared_breaks
)

df_long_cv1 <- compute_diff_per_interval(
  all_stats_real, all_stats,
  metric = "log_cv",
  num_intervals = 25,
  group_index = 2,
  breaks = shared_breaks
)

df_long_cv$group  <- "AAM"
df_long_cv1$group <- "AFR"

df_ridge_cv <- rbind(df_long_cv, df_long_cv1)


bridge_CV <- ggplot(df_ridge_cv, aes(x = diff, y = interval_label, fill = group)) +
  geom_density_ridges(alpha = 0.7) +
  scale_fill_viridis_d(option = "D", end = 0.9) +
  theme_ridges() +
  theme(
    # Smaller y-axis text
    axis.text.y = element_text(size = 7),       
    axis.title.y = element_text(size = 8, face = "bold", hjust = 0.5),
    
    # Smaller x-axis text
    axis.text.x = element_text(size = 8),
    axis.title.x = element_text(size = 10, face = "bold", hjust = 0.5),
    
    # Smaller legend text
    legend.title = element_text(size = 7, face = "bold"),
    legend.text  = element_text(size = 6),
    
    # Adjust strip text (facet labels)
    strip.text = element_text(size = 9, face = "bold"),
    
    # Caption formatting: smaller font and centered
    plot.caption = element_text(size = 10, hjust = 0.5, lineheight = 1.2),
    
    # Increase spacing between facets if needed
    panel.spacing = unit(1, "lines")
  ) +
  labs(
    x = expression("Simulated – Real ("*Delta*" mean log-CV)"),
    y = "Intervals",
    caption = paste0(
      "Figure 4c: Distribution of differences between simulated and real mean log-CV\n",
      "across 24 mean-abundance intervals (minimum 5 taxa per interval),\n",
      "based on 100 simulations for for both AAM and AFR."
    ),
    fill = "Group"
  ) +
  facet_grid(. ~ group)

```

```{r , echo=FALSE, warning = FALSE, message = FALSE}

st1 <- as.data.frame(get_stats(otu1, rownames(meta1))) %>%
  mutate(source = "Simulated Data")

# Real data
st2 <- as.data.frame(get_stats(otu_real, rownames(meta_real))) %>%
  mutate(source = "Real Data")

# Combine
df <- bind_rows(st1, st2)

# --- Plot 1: variance vs mean ---
plot_var <- ggplot(df, aes(x = log_mean, y = log_var, color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "var log2-CPM",
    caption = "(a)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )

# --- Plot 2: zeros vs mean ---
plot_zero <- ggplot(df, aes(x = log_mean, y = frac_zero, color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "fraction zeros",
    title = "(a)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )

# --- Plot 3: cv vs mean ---
plot_cv <- ggplot(df, aes(x = log_mean, y = log_cv , color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "log2-CV",
    caption = "(a)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )
```

```{r , echo=FALSE, warning = FALSE, message = FALSE}
# --- Compute stats for ALL simulated samples ---
samples_sim <- rownames(meta_mid_nonpar)
st_sim <- as.data.frame(get_stats(otu_sim, samples_sim)) %>%
  mutate(source = "Simulated Data")

# --- Compute stats for ALL real samples ---
samples_real <- rownames(meta_real)
st_real <- as.data.frame(get_stats(otu_real_non, samples_real)) %>%
  mutate(source = "Real Data")

# Combine
df <- bind_rows(st_sim, st_real)

# --- Plot 1: variance vs mean ---
plot_var_non <- ggplot(df, aes(x = log_mean, y = log_var, color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "var log2-CPM",
    caption = "(b)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )

# --- Plot 2: zeros vs mean ---
plot_zero_non <- ggplot(df, aes(x = log_mean, y = frac_zero, color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "fraction zeros",
    caption  = "(b)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )

# --- Plot 3: CV vs mean ---
plot_cv_non <- ggplot(df, aes(x = log_mean, y = log_cv, color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "log2-CV",
    caption = "(b)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )

```

```{r , echo=FALSE, warning = FALSE, message = FALSE}
# --- Compute stats for ALL SPsimSeq samples ---
samples_sim <- rownames(meta_spsim)
st_sim <- as.data.frame(get_stats(otu_spsim, samples_sim)) %>%
  mutate(source = "SPsimSeq")

# --- Compute stats for ALL real samples ---
samples_real <- rownames(meta_real)
st_real <- as.data.frame(get_stats(otu_real_filtered, samples_real)) %>%
  mutate(source = "Real Data")

# --- Combine ---
df <- bind_rows(st_sim, st_real)

# --- Plot 1: variance vs mean ---
plot_var_spsim <- ggplot(df, aes(x = log_mean, y = log_var, color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "var log2-CPM",
    caption  = "(c)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )

# --- Plot 2: zeros vs mean ---
plot_zero_spsim <- ggplot(df, aes(x = log_mean, y = frac_zero, color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "fraction zeros",
    caption  = "(c)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )

# --- Plot 3: CV vs mean ---
plot_cv_spsim <- ggplot(df, aes(x = log_mean, y = log_cv, color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "log2-CV",
    caption  = "(c)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )
```

```{r , echo=FALSE, warning = FALSE, message = FALSE}
# --- Compute stats for ALL SparseDOSSA2 simulated samples ---
samples_sim <- rownames(meta_Spars)
st_sim <- as.data.frame(get_stats(otu_sim2, samples_sim)) %>%
  mutate(source = "Simulated Data")

# --- Compute stats for ALL real samples ---
samples_real <- rownames(meta_real)
st_real <- as.data.frame(get_stats(otu_real_filtered, samples_real)) %>%
  mutate(source = "Real Data")

# Combine
df <- bind_rows(st_sim, st_real)

# --- Plot 1: variance vs mean ---
plot_var_spars <- ggplot(df, aes(x = log_mean, y = log_var, color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "var log2-CPM",
    caption = "(d)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )

# --- Plot 2: zeros vs mean ---
plot_zero_spars <- ggplot(df, aes(x = log_mean, y = frac_zero, color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "fraction zeros",
    caption  = "(d)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )

# --- Plot 3: CV vs mean ---
plot_cv_spars <- ggplot(df, aes(x = log_mean, y = log_cv, color = source)) +
  geom_point(alpha = 1, size = 0.2) +
  labs(
    x = "mean log2-CPM",
    y = "log2_CV",
    caption  = "(d)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )

```


```{r , echo=FALSE, warning = FALSE, message = FALSE}

combined1 <- (plot_var | plot_var_non) /
  (plot_var_spsim | plot_var_spars) +
  plot_annotation(
    caption = paste0(
      "Figure 5a: Comparison of the variance–mean relationship for taxon abundances between the real\n",
      " dataset and single simulated datasets across different simulation methods\n",
      "a: MIDASim (parametric), b: MIDASim (non-parametric), c: SPsimSeq, d: SparseDOSSA2"
    )
  ) &
  theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))
```
```{r, fig.height=4, fig.width=6, out.height="45%", out.width="90%", echo=FALSE, warning = FALSE, message = FALSE}
bridge_CV
combined1
```


```{r , echo=FALSE, warning = FALSE, message = FALSE}

combined3 <- (plot_cv | plot_cv_non) /
  (plot_cv_spsim | plot_cv_spars) +
  plot_annotation(
    caption = paste0(
      "Figure 5b: Comparison of the CV–mean relationship for taxon abundances between the real\n",
      " dataset and single simulated datasets across different simulation methods\n",
      "a: MIDASim (parametric), b: MIDASim (non-parametric), c: SPsimSeq, d: SparseDOSSA2")
    ) &
  theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))
```

```{r , echo=FALSE, warning = FALSE, message = FALSE}

combined5 <- (plot_zero | plot_zero_non) /
  (plot_zero_spsim | plot_zero_spars) +
  plot_annotation(
    caption = paste0(
      "Figure 5c: Comparison of the fraction of zeros vs mean relationship for taxon abundances between the real\n",
      " dataset and single simulated datasets across different simulation methods\n",
      "a: MIDASim (parametric), b: MIDASim (non-parametric), c: SPsimSeq, d: SparseDOSSA2")
  ) &
  theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))
```
```{r, fig.height=4, fig.width=6, out.height="45%", out.width="90%", echo=FALSE, warning = FALSE, message = FALSE}
combined5
combined3
```

# 4. Conclusions

Both single and 100 parametric-mode simulations closely resembled the template data, particularly reproducing the mean–variance structure; however, the simulator was less accurate at capturing sparsity. Non-parametric simulations performed better in replicating the real data patterns and make relatively few distributional assumptions about relative abundances. Nonetheless, non-parametric methods require a minimum number of non-zero counts per taxon to estimate distributions reliably, which limited the number of simulated samples compared to the original dataset.

# 4. Reference

O’Keefe, S. J., J. V. Li, L. Lahti, J. Ou, F. Carbonero, K. Mohammed, J. M. Posma, J. Kinross, E. Wahl and E. Ruder (2015). "Fat, fibre and cancer risk in African Americans and rural Africans." Nature communications 6(1): 6342.

Assefa, A. T., J. Vandesompele and O. Thas (2020). "SPsimSeq: semi-parametric simulation of bulk and single-cell RNA-sequencing data." Bioinformatics 36(10): 3276-3278.

He, M., N. Zhao and G. A. Satten (2024). "MIDASim: a fast and simple simulator for realistic microbiome data." Microbiome 12(1): 135.

Assefa (2022) provided the SPsimSeq code on GitHub (https://github.com/CenterForStatistics-UGent/SPsimSeq/blob/master/R/SPsimSeq.R, accessed 2025-12-05).

Ma (2021) provided the SparseDOSSA2 code on GitHub (https://github.com/biobakery/SparseDOSSA2/blob/master/R/SparseDOSSA2_fit.R, accessed 2025-12-05).

He (2025) provided the MIDASim code on GitHub (https://github.com/mengyu-he/MIDASim/tree/master/R, accessed 2025-12-05).
